---
title: "Coordinate exchange compared with point exchange"
format:
  html:
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Coordinate exchange compared with point exchange}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: false
  warning: false
---

This benchmark compares the public `gen_design()` workflows and the underlying
search kernels separately. Public timings include setup and method-native random
starts. Kernel timings exclude shared setup and give both methods the same
feasible initial designs.

```{r}
result_dir = file.path("inst", "benchmarks", "results")
if (!dir.exists(result_dir)) result_dir = file.path("..", "inst", "benchmarks", "results")
summary_path = file.path(result_dir, "ce_point_smoke_summary.csv")
paired_path = file.path(result_dir, "ce_point_smoke_paired.csv")
if (!file.exists(summary_path) || !file.exists(paired_path)) {
  stop(
    "Run inst/benchmarks/run_coordinate_exchange_benchmarks.R ",
    "with --profile=smoke before rendering this report."
  )
}
summary = read.csv(summary_path, stringsAsFactors = FALSE)
paired = read.csv(paired_path, stringsAsFactors = FALSE)
```

## Quality and correctness

Every successful coordinate-exchange result must be feasible, finite, and full
rank. The automated gate requires every paired D-efficiency ratio to be at least
`0.95` and the overall median to be at least `0.99`.

```{r}
knitr::kable(
  summary[, c(
    "track", "scenario", "successful_pairs", "failure_rate",
    "d_ratio_median", "d_ratio_low", "d_ratio_high"
  )],
  digits = 4,
  caption = "Coordinate/point D-efficiency ratios"
)
```

## Runtime

Runtime is reported as point exchange divided by coordinate exchange. Values
above one favor coordinate exchange. Intervals are seeded percentile-bootstrap
intervals over paired runs, not model-based confidence claims.

```{r}
knitr::kable(
  summary[, c(
    "track", "scenario", "runtime_ratio_median", "runtime_ratio_iqr",
    "runtime_ratio_low", "runtime_ratio_high"
  )],
  digits = 3,
  caption = "Paired point/coordinate runtime ratios"
)
```

The scenario matrix includes mixed factors, interactions, coupled constraints,
forbidden tuples, disconnected and sparse spaces, oversized groups, masked
formula dependencies, near-saturated models, and ill-conditioned models. See
`inst/benchmarks/README.md` for implementation diagrams and exact commands.
