---
title: "CE vs Point: Large Design-Space Experiments"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
---

## Overview

This report runs a broad benchmark comparison between point-exchange (`fedorov`) and coordinate-exchange across:

- `trial_size`: `small`, `large`
- `candidate_size`: `small`, `large`
- interaction complexity: `none`, `some`, `many`
- quadratic complexity: `none`, `some`, `many`
- nominal levels: `3`, `4`, `5`
- numeric density: `low` (`-1:1 by 1`) vs `high` (`-1:1 by 0.1`)

For each selected scenario, both methods are run with matched random seeds over multiple random iterations.

A power-driven meta-design is first generated using `skpr` to target:

- `alpha = 0.05`
- `effectsize (SNR) = 0.5`
- `power >= 0.95` for the `method` effect.

```{r}
library(data.table)
library(ggplot2)
library(knitr)
```

## Generate Data

```{r}
resolve_path <- function(candidates) {
  hit <- candidates[file.exists(candidates)]
  if (length(hit) == 0) {
    return(candidates[[1]])
  }
  hit[[1]]
}

run_script <- resolve_path(c(
  "run_ce_point_designspace_experiments.R",
  "inst/benchmarks/run_ce_point_designspace_experiments.R"
))

power_scan_path <- resolve_path(c(
  "results/designspace_power_scan.csv",
  "inst/benchmarks/results/designspace_power_scan.csv"
))
raw_path <- resolve_path(c(
  "results/designspace_experiment_raw.csv",
  "inst/benchmarks/results/designspace_experiment_raw.csv"
))
paired_path <- resolve_path(c(
  "results/designspace_experiment_paired.csv",
  "inst/benchmarks/results/designspace_experiment_paired.csv"
))
lm_analysis_script <- resolve_path(c(
  "analyze_designspace_with_lm.R",
  "inst/benchmarks/analyze_designspace_with_lm.R"
))
lm_analysis_path <- resolve_path(c(
  "results/designspace_lm_analysis.txt",
  "inst/benchmarks/results/designspace_lm_analysis.txt"
))
lm_quality_global_coef_path <- resolve_path(c(
  "results/designspace_lm_quality_global_coefficients.csv",
  "inst/benchmarks/results/designspace_lm_quality_global_coefficients.csv"
))
lm_quality_coef_path <- resolve_path(c(
  "results/designspace_lm_quality_coefficients.csv",
  "inst/benchmarks/results/designspace_lm_quality_coefficients.csv"
))
lm_quality_anova_path <- resolve_path(c(
  "results/designspace_lm_quality_anova.csv",
  "inst/benchmarks/results/designspace_lm_quality_anova.csv"
))
lm_runtime_coef_path <- resolve_path(c(
  "results/designspace_lm_runtime_coefficients.csv",
  "inst/benchmarks/results/designspace_lm_runtime_coefficients.csv"
))
lm_runtime_anova_path <- resolve_path(c(
  "results/designspace_lm_runtime_anova.csv",
  "inst/benchmarks/results/designspace_lm_runtime_anova.csv"
))

if (!file.exists(power_scan_path) || !file.exists(raw_path) || !file.exists(paired_path)) {
  system2(
    command = file.path(R.home("bin"), "Rscript"),
    args = run_script,
    stdout = TRUE,
    stderr = TRUE
  )
}

if (!file.exists(lm_analysis_path) || !file.exists(lm_quality_coef_path)) {
  system2(
    command = file.path(R.home("bin"), "Rscript"),
    args = lm_analysis_script,
    stdout = TRUE,
    stderr = TRUE
  )
}

power_scan <- fread(power_scan_path)
raw <- fread(raw_path)
paired <- fread(paired_path)
lm_stats_txt <- if (file.exists(lm_analysis_path)) readLines(lm_analysis_path) else character(0)
lm_quality_global_coef <- fread(lm_quality_global_coef_path)
lm_quality_coef <- fread(lm_quality_coef_path)
lm_quality_anova <- fread(lm_quality_anova_path)
lm_runtime_coef <- fread(lm_runtime_coef_path)
lm_runtime_anova <- fread(lm_runtime_anova_path)
```

## Power-Design Scan

```{r}
kable(power_scan[order(trials)], digits = 4, caption = "Method-effect power scan (SNR=0.5, alpha=0.05)")
```

```{r}
ggplot(power_scan[ok == TRUE], aes(x = trials, y = method_power)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  labs(
    title = "Method-effect power vs meta-design trial count",
    x = "Meta-design trials",
    y = "Power for method effect"
  ) +
  theme_minimal()
```

## Run Summary

```{r}
raw[, method := as.character(method)]
paired[, both_ok := as.logical(both_ok)]

summary_table <- data.table(
  total_runs = nrow(raw),
  successful_runs = raw[is.na(error), .N],
  paired_rows = nrow(paired),
  paired_success = paired[both_ok == TRUE, .N]
)

kable(summary_table, caption = "Execution summary")
```

## LM Analysis (Primary)

```{r}
if (length(lm_stats_txt) > 0) {
  cat(paste(lm_stats_txt, collapse = "\n"))
}
```

```{r}
kable(
  lm_quality_global_coef,
  digits = 6,
  caption = "Global paired-difference lm: d_opt_diff_ce_minus_point ~ 1"
)
```

```{r}
kable(
  lm_quality_coef,
  digits = 6,
  caption = "Adjusted lm coefficients: d_opt_diff_ce_minus_point ~ trial_size + candidate_size + interactions + quadratic + nominal_levels + density"
)
```

```{r}
kable(
  lm_quality_anova,
  digits = 6,
  caption = "Adjusted lm ANOVA: D-optimality difference"
)
```

```{r}
kable(
  lm_runtime_coef,
  digits = 6,
  caption = "Adjusted lm coefficients: log(runtime_ce/runtime_point) ~ factors"
)
```

```{r}
kable(
  lm_runtime_anova,
  digits = 6,
  caption = "Adjusted lm ANOVA: Runtime log ratio"
)
```

## D-Optimality Difference (CE - Point)

```{r}
paired_ok <- paired[both_ok == TRUE & is.finite(d_opt_diff_ce_minus_point)]

kable(
  paired_ok[, .(
    n = .N,
    mean_diff = mean(d_opt_diff_ce_minus_point),
    median_diff = median(d_opt_diff_ce_minus_point),
    sd_diff = sd(d_opt_diff_ce_minus_point)
  )],
  digits = 6,
  caption = "Overall CE - Point D-optimality difference"
)
```

```{r}
ggplot(paired_ok, aes(x = d_opt_diff_ce_minus_point)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of D-opt difference (CE - Point)",
    x = "D-opt difference",
    y = "Count"
  ) +
  theme_minimal()
```

## Runtime Ratio (Point / CE)

```{r}
kable(
  paired_ok[, .(
    n = .N,
    mean_point_over_ce = mean(runtime_ratio_point_over_ce, na.rm = TRUE),
    median_point_over_ce = median(runtime_ratio_point_over_ce, na.rm = TRUE),
    p90_point_over_ce = quantile(runtime_ratio_point_over_ce, probs = 0.9, na.rm = TRUE)
  )],
  digits = 4,
  caption = "Runtime ratio summary (point runtime / CE runtime)"
)
```

## Factor-Wise Effects

```{r}
factor_cols <- c("trial_size", "candidate_size", "interactions", "quadratic", "nominal_levels", "density")

for (fc in factor_cols) {
  cat("\n### ", fc, "\n", sep = "")
  tab <- paired_ok[, .(
    n = .N,
    mean_diff = mean(d_opt_diff_ce_minus_point),
    median_diff = median(d_opt_diff_ce_minus_point),
    sd_diff = sd(d_opt_diff_ce_minus_point),
    mean_runtime_ratio = mean(runtime_ratio_point_over_ce, na.rm = TRUE)
  ), by = fc][order(get(fc))]
  print(kable(tab, digits = 5))
}
```

## Raw Data Files

Outputs are written to:

- `inst/benchmarks/results/designspace_power_scan.csv`
- `inst/benchmarks/results/designspace_powered_meta_design.csv`
- `inst/benchmarks/results/designspace_base_scenarios.csv`
- `inst/benchmarks/results/designspace_experiment_raw.csv`
- `inst/benchmarks/results/designspace_experiment_paired.csv`
- `inst/benchmarks/results/designspace_experiment_summary_by_scenario.csv`
- `inst/benchmarks/results/designspace_experiment_stats.txt`
