---
title: "Coordinate-Exchange vs Point-Exchange Benchmarks"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
---

## Overview

This document benchmarks `skpr::gen_design()` using:

- `advancedoptions$search_method = "fedorov"` (point-exchange)
- `advancedoptions$search_method = "coordinate_exchange"` (CE)

The benchmark set includes:

1. Prior size/trial matrix (candidate set size and design size combinations)
2. Scaling with respect to `trials`
3. Scaling with respect to candidate-set rows (the CE search space rows)
4. Scaling with respect to constraints (feasible rows after applying constraints)
5. Scaling with respect to model complexity (model-matrix columns)

All timings use the `microbenchmark` package.

The current sweep extends into larger regimes to reduce constant-term dominance:

- trial counts up to `3000`
- candidate sets up to approximately `30,000` rows

```{r}
library(data.table)
library(ggplot2)
library(knitr)
```

## Generate Benchmark Data

```{r}
resolve_path <- function(candidates) {
  hit <- candidates[file.exists(candidates)]
  if (length(hit) == 0) {
    return(candidates[[1]])
  }
  hit[[1]]
}

benchmark_script <- resolve_path(c("run_ce_point_benchmarks.R", "inst/benchmarks/run_ce_point_benchmarks.R"))
quality_script <- resolve_path(c("run_ce_point_quality.R", "inst/benchmarks/run_ce_point_quality.R"))
summary_path <- resolve_path(c("results/benchmark_summary.csv", "inst/benchmarks/results/benchmark_summary.csv"))
raw_path <- resolve_path(c("results/benchmark_raw.csv", "inst/benchmarks/results/benchmark_raw.csv"))
speed_path <- resolve_path(c("results/benchmark_speedup.csv", "inst/benchmarks/results/benchmark_speedup.csv"))
scaling_path <- resolve_path(c("results/scaling_exponents.csv", "inst/benchmarks/results/scaling_exponents.csv"))
quality_path <- resolve_path(c("results/benchmark_quality.csv", "inst/benchmarks/results/benchmark_quality.csv"))
quality_compare_path <- resolve_path(c("results/benchmark_quality_compare.csv", "inst/benchmarks/results/benchmark_quality_compare.csv"))

if (!file.exists(summary_path) || !file.exists(raw_path) ||
    !file.exists(speed_path) || !file.exists(scaling_path)) {
  system2(
    command = file.path(R.home("bin"), "Rscript"),
    args = benchmark_script,
    stdout = TRUE,
    stderr = TRUE
  )
}

if (!file.exists(quality_path) || !file.exists(quality_compare_path)) {
  system2(
    command = file.path(R.home("bin"), "Rscript"),
    args = quality_script,
    stdout = TRUE,
    stderr = TRUE
  )
}
```

```{r}
bench_summary <- fread(summary_path)
bench_raw <- fread(raw_path)
bench_speed <- fread(speed_path)
scaling <- fread(scaling_path)
quality <- fread(quality_path)
quality_compare <- fread(quality_compare_path)

bench_summary[, expr := as.character(expr)]
bench_summary[, section := as.character(section)]
bench_summary[, scenario := as.character(scenario)]
bench_speed[, section := as.character(section)]
bench_speed[, scenario := as.character(scenario)]
quality[, section := as.character(section)]
quality[, scenario := as.character(scenario)]
quality_compare[, section := as.character(section)]
quality_compare[, scenario := as.character(scenario)]
```

## Prior Benchmark Matrix

This section reproduces the prior benchmark matrix (candidate set size x trial count).

```{r}
prior <- bench_summary[section == "prior"][
  order(candidate_rows, trials, expr),
  .(scenario, expr, candidate_rows, trials, repeats, min, median, mean, max)
]
kable(prior, digits = 2, caption = "Prior benchmark matrix (milliseconds)")
```

```{r}
prior_speed <- bench_speed[section == "prior"][
  order(candidate_rows, trials),
  .(scenario, candidate_rows, trials, median_point, median_ce, point_over_ce)
]
kable(prior_speed, digits = 3, caption = "Prior matrix speed ratio: point/CE (median)")
```

## Scaling Fits

Scaling is estimated with log-log linear models:

`log(time_ms) = alpha + beta * log(size_variable)`

Estimated `beta` is interpreted as the empirical O(n^beta) exponent.

```{r}
kable(
  scaling[order(section, method), .(section, xvar, method, exponent, r_squared, n)],
  digits = 4,
  caption = "Empirical scaling exponents (log-log fits)"
)
```

## 1) Scaling w.r.t. Trials

```{r}
d_trials <- bench_summary[section == "scaling_trials"]

ggplot(d_trials, aes(x = trials, y = median, color = expr)) +
  geom_point(size = 2) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  labs(
    title = "Median time vs trials",
    x = "Trials (log scale)",
    y = "Median runtime ms (log scale)",
    color = "Method"
  ) +
  theme_minimal()
```

```{r}
kable(
  d_trials[order(trials, expr), .(scenario, expr, trials, median, mean, max)],
  digits = 2,
  caption = "Trials scaling raw summary"
)
```

## 2) Scaling w.r.t. Candidate Rows

Here, candidate set size is varied while keeping the model and trial count fixed.

```{r}
d_rows <- bench_summary[section == "scaling_candidate_rows"]

ggplot(d_rows, aes(x = candidate_rows, y = median, color = expr)) +
  geom_point(size = 2) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  labs(
    title = "Median time vs candidate rows",
    x = "Candidate rows (log scale)",
    y = "Median runtime ms (log scale)",
    color = "Method"
  ) +
  theme_minimal()
```

```{r}
kable(
  d_rows[order(candidate_rows, expr), .(scenario, expr, candidate_rows, trials, median)],
  digits = 2,
  caption = "Candidate-row scaling raw summary"
)
```

## 3) Scaling w.r.t. Constraints

Constraint intensity is represented by `feasible_rows` (rows remaining after constraint filtering).

For point-exchange, reducing feasible rows reduces the candidate list directly, which typically speeds search.

```{r}
d_constraints <- bench_summary[section == "scaling_constraints"]

ggplot(d_constraints, aes(x = feasible_rows, y = median, color = expr)) +
  geom_point(size = 2) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  labs(
    title = "Median time vs feasible rows under constraints",
    x = "Feasible rows (log scale)",
    y = "Median runtime ms (log scale)",
    color = "Method"
  ) +
  theme_minimal()
```

```{r}
kable(
  d_constraints[
    order(feasible_rows, expr),
    .(scenario, constraints_label, expr, feasible_rows, feasible_fraction, median, mean, max)
  ],
  digits = 3,
  caption = "Constraints scaling raw summary"
)
```

## 4) Scaling w.r.t. Model Complexity

Model complexity is proxied by `complexity_cols = ncol(model.matrix(...))`.

```{r}
d_model <- bench_summary[section == "scaling_model_complexity"]

ggplot(d_model, aes(x = complexity_cols, y = median, color = expr)) +
  geom_point(size = 2) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  labs(
    title = "Median time vs model-matrix columns",
    x = "Model-matrix columns (log scale)",
    y = "Median runtime ms (log scale)",
    color = "Method"
  ) +
  theme_minimal()
```

```{r}
kable(
  d_model[order(complexity_cols, expr), .(scenario, model_label, expr, complexity_cols, median, mean, max)],
  digits = 3,
  caption = "Model complexity scaling raw summary"
)
```

## Consolidated Speed Ratios

`point_over_ce > 1` means CE is faster.  
`point_over_ce < 1` means point-exchange is faster.

```{r}
kable(
  bench_speed[
    order(section, scenario),
    .(section, scenario, candidate_rows, trials, constraints_label, feasible_rows, complexity_cols, median_point, median_ce, point_over_ce)
  ],
  digits = 3,
  caption = "Consolidated speed-ratio table"
)
```

## Design Quality (D-Optimality)

```{r}
kable(
  quality_compare[
    order(section, scenario),
    .(section, scenario, candidate_rows, trials, feasible_rows, complexity_cols,
      d_opt_point, d_opt_ce, d_opt_diff_ce_minus_point, d_opt_ratio_ce_over_point)
  ],
  digits = 4,
  caption = "D-optimality comparison (point vs CE)"
)
```

```{r}
quality_trials <- quality[section == "scaling_trials"]

ggplot(quality_trials, aes(x = trials, y = d_opt, color = expr)) +
  geom_point(size = 2) +
  geom_line() +
  scale_x_log10() +
  labs(
    title = "D-optimality vs trials",
    x = "Trials (log scale)",
    y = "D-efficiency",
    color = "Method"
  ) +
  theme_minimal()
```

## Files Written

Benchmark data files:

- `inst/benchmarks/results/benchmark_summary.csv`
- `inst/benchmarks/results/benchmark_raw.csv`
- `inst/benchmarks/results/benchmark_speedup.csv`
- `inst/benchmarks/results/scaling_exponents.csv`
- `inst/benchmarks/results/benchmark_quality.csv`
- `inst/benchmarks/results/benchmark_quality_compare.csv`
